---
title: "Research Variables Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This document analyzes the variables available in `R Raw Data.csv`:

- **Tiempo (sem)** — time in weeks
- **¿Llegó a término?** — terminal endpoint indicator (`1` = Yes, `0` = No)
- **Sexo** — sex category (`Macho` / `Hembra`)

---

## 1. Load Libraries and Data

```{r load}
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(scales)

# Load and normalize CSV columns
raw_data <- read.csv("R Raw Data.csv", stringsAsFactors = FALSE, skip = 1, check.names = FALSE)

data <- raw_data %>%
  transmute(
    weeks_at_euthanasia = as.numeric(sub(",", ".", `Tiempo (sem)`, fixed = TRUE)),
    reached_terminal = ifelse(trimws(as.character(`¿Llegó a término?`)) == "1", "Yes", "No"),
    sex = case_when(
      tolower(trimws(Sexo)) == "macho"  ~ "M",
      tolower(trimws(Sexo)) == "hembra" ~ "F",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    age_weeks = weeks_at_euthanasia,
    sex = factor(sex, levels = c("M", "F"), labels = c("Male", "Female")),
    reached_terminal = factor(reached_terminal, levels = c("Yes", "No"))
  )

kable(head(data), caption = "First rows of the dataset")
```

---

## 2. Tiempo (sem)

```{r age-summary}
cat("Summary of Tiempo (sem):\n")
summary(data$age_weeks)
```

```{r age-plot}
ggplot(data, aes(x = factor(age_weeks))) +
  geom_bar(fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Tiempo (sem)",
    x     = "Tiempo (sem)",
    y     = "Count"
  ) +
  theme_minimal()
```

---

## 3. Sex

```{r sex-summary}
sex_table <- table(data$sex)
kable(as.data.frame(sex_table), col.names = c("Sex", "Count"),
      caption = "Animal counts by sex")
```

```{r sex-plot}
ggplot(data, aes(x = sex, fill = sex)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_manual(values = c("Male" = "#4E79A7", "Female" = "#F28E2B")) +
  labs(
    title = "Sex Distribution",
    x     = "Sex",
    y     = "Count"
  ) +
  theme_minimal()
```

---

## 4. Weeks at Euthanasia

```{r euthanasia-summary}
cat("Summary of Weeks at Euthanasia:\n")
summary(data$weeks_at_euthanasia)
```

```{r euthanasia-plot}
ggplot(data, aes(x = weeks_at_euthanasia, fill = sex)) +
  geom_histogram(binwidth = 2, color = "white", position = "dodge") +
  scale_fill_manual(values = c("Male" = "#4E79A7", "Female" = "#F28E2B")) +
  labs(
    title = "Distribution of Age at Euthanasia by Sex",
    x     = "Weeks at Euthanasia",
    y     = "Count",
    fill  = "Sex"
  ) +
  theme_minimal()
```

```{r euthanasia-boxplot}
ggplot(data, aes(x = sex, y = weeks_at_euthanasia, fill = sex)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = 21) +
  scale_fill_manual(values = c("Male" = "#4E79A7", "Female" = "#F28E2B")) +
  labs(
    title = "Age at Euthanasia by Sex",
    x     = "Sex",
    y     = "Weeks at Euthanasia"
  ) +
  theme_minimal()
```

---

## 5. Terminal Endpoint (Tumor Growth)

```{r terminal-summary}
terminal_table <- table(data$reached_terminal)
terminal_pct   <- prop.table(terminal_table) * 100

kable(
  data.frame(
    Reached_Terminal = names(terminal_table),
    Count            = as.integer(terminal_table),
    Percent          = round(as.numeric(terminal_pct), 1)
  ),
  caption = "Terminal endpoint counts and percentages"
)
```

```{r terminal-plot}
ggplot(data, aes(x = reached_terminal, fill = reached_terminal)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_manual(values = c("Yes" = "#E15759", "No" = "#76B7B2")) +
  labs(
    title = "Animals Reaching Terminal Endpoint",
    x     = "Reached Terminal (Tumor Growth)",
    y     = "Count"
  ) +
  theme_minimal()
```

---

## 6. Cross-tabulations

### Terminal endpoint by sex

```{r crosstab-sex-terminal}
ct <- table(Sex = data$sex, Terminal = data$reached_terminal)
kable(ct, caption = "Terminal endpoint by sex")
```

```{r crosstab-sex-terminal-plot}
ggplot(data, aes(x = sex, fill = reached_terminal)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Yes" = "#E15759", "No" = "#76B7B2")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Proportion Reaching Terminal Endpoint by Sex",
    x     = "Sex",
    y     = "Proportion",
    fill  = "Reached Terminal"
  ) +
  theme_minimal()
```

### Weeks at euthanasia by terminal endpoint

```{r euthanasia-by-terminal}
data %>%
  group_by(reached_terminal) %>%
  summarise(
    n        = n(),
    mean_wks = round(mean(weeks_at_euthanasia), 1),
    sd_wks   = round(sd(weeks_at_euthanasia), 1),
    min_wks  = min(weeks_at_euthanasia),
    max_wks  = max(weeks_at_euthanasia)
  ) %>%
  kable(
    col.names = c("Reached Terminal", "N", "Mean Weeks", "SD", "Min", "Max"),
    caption   = "Weeks at euthanasia by terminal endpoint"
  )
```

```{r euthanasia-terminal-boxplot}
ggplot(data, aes(x = reached_terminal, y = weeks_at_euthanasia, fill = reached_terminal)) +
  geom_boxplot(show.legend = FALSE, outlier.shape = 21) +
  scale_fill_manual(values = c("Yes" = "#E15759", "No" = "#76B7B2")) +
  labs(
    title = "Weeks at Euthanasia by Terminal Endpoint",
    x     = "Reached Terminal",
    y     = "Weeks at Euthanasia"
  ) +
  theme_minimal()
```

---

## 7. Statistical Tests

### Difference in weeks at euthanasia: terminal vs. non-terminal

```{r t-test}
t_result <- t.test(weeks_at_euthanasia ~ reached_terminal, data = data)
cat(sprintf(
  "Welch two-sample t-test\n  t = %.3f, df = %.1f, p-value = %.4f\n",
  t_result$statistic, t_result$parameter, t_result$p.value
))
```

### Association between sex and terminal endpoint

```{r chi-sq}
chi_result <- chisq.test(table(data$sex, data$reached_terminal))
cat(sprintf(
  "Chi-squared test\n  X² = %.3f, df = %d, p-value = %.4f\n",
  chi_result$statistic, chi_result$parameter, chi_result$p.value
))
```

---

## 8. Session Info

```{r session-info}
sessionInfo()
```
